---
title: "Shotgun Workshop - Downstream analysis with R"
author: "Hena R. Ramay & Ewandson L. Lameu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    theme: yeti
    highlight: kate
    toc: true
    toc_float: 
      collapsed: false
    toc_depth: 4
editor_options: 
  chunk_output_type: inline
keep_md: no
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,message = F,warning = F,fig.height = 5,fig.width = 7, include = T)
```

# **Loading, editing and converting data to a Phyloseq Object**

Before running the following R code, please make sure that you have downloaded the complete example folder with data folder inside it.

 <br>
 
### Load Libraries
```{r libraries, echo=T}
library(tidyverse)
library(phyloseq)


```

<br>

### Setup data files 
```{r 2, echo=T}
# load otu table
otu <- read_delim(file="data/merged_raw_counts_species.txt") %>% 
                  data.frame() %>%
                  filter(clade_name!="UNCLASSIFIED")  # needs to be a dataframe
                                                                                         
rownames(otu)<-otu$clade_name  # add row names

#create taxa table
taxa <- otu %>% select(clade_name) %>% separate(col=clade_name, 
                                                into = c("Kingdom",
                                                         "Phylum",
                                                         "Class",
                                                         "Order",
                                                         "Family",
                                                         "Genus",
                                                         "Species"),
                                                sep="[|]") %>% 
                                      as.matrix()  # needs to be a matrix

otu<-otu %>% select(-clade_name)  # delete the clade_name column   
colnames(otu)<-str_remove_all(string=colnames(otu),
                              pattern="_metaphlan") # format column names 

# load sample info (metadata) table
sample_info<-read_delim(file="data/sample_info.csv") %>% 
                        data.frame() # needs to be a dataframe
rownames(sample_info)<-sample_info$Sample  # add row names 

```

<br>

### Create a Phyloseq Object
```{r phyloseq object}
# create phyloseq object
ps<- phyloseq(otu_table(otu, taxa_are_rows = T), 
              tax_table(taxa),
              sample_data(sample_info)) 

```

<br>

### Prune Samples by total number of reads
```{r filtering samples and taxa based on number of reads}

# filtering samples
colSums(ps@otu_table) # check sample reads

# remove samples S8 with no reads
samples_to_keep<- ps@sam_data$Sample[ps@sam_data$Sample!="S8"]
ps <- prune_samples(samples = samples_to_keep, ps)  # now S8 was removed from ps

# Another way of doing this is
# ps@sam_data<-ps@sam_data %>% data.frame() %>% filter(Sample !="S8") %>% sample_data()


```
<br>

### Prune Taxa by total number of reads
```{r filtering taxa based on number of reads}

# filtering taxa
unname(taxa_sums(ps)) # check taxa reads, unname removes the taxa names as they are very long.
ps<- prune_taxa(taxa_sums(ps)>5000, ps)    # remove taxa with less than 5000 reads, 
                                           # please use appropriate value specific to your dataset


```

<br>

### Create relative abundance and Top 5 taxa files

```{r relative abundance and top5 taxas}
# relative abundance table
ps_rel_abun <- ps %>% transform_sample_counts(function(x) x/sum(x))

# transformed data, to be used for beta diversity. 
#You can use other transformations too for beta diversity like VST,CLR etc
ps_log <- ps %>% transform_sample_counts(function(x) log(x+1))

# Prune only the top 5 taxas

top5 <- names(sort(taxa_sums(ps), decreasing = T))[1:5]   # names of top 5 taxas
ps_top5 <-prune_taxa(taxa=top5, ps) # prune to keep only the top 5 on the phyloseq object

```

<br>

# **Abundance Plots**

```{r abundance plots}

#abundance plots for count reads
plot_bar(physeq=ps, fill="Phylum")

#abundance plots for relative abundances
plot_bar(physeq=ps_rel_abun, fill="Phylum")

#abundance plots for count reads over the top 5 taxa only
plot_bar(physeq=ps_top5, fill="Phylum")


```

<br>

# **Alpha Diversity**

```{r alpha diversity}

# plot alpha diversity for the chosen distance metric
plot_richness(ps, x ="Type", measures = c("Shannon","Observed","Simpson"),
              color = "Type")

# plot alpha diversity in boxplot format for the chosen distance metric
plot_richness(ps, x ="Type", measures = c("Shannon","Observed","Simpson"),
              color = "Type") +
              geom_boxplot()

# save richness in a table
richness <- estimate_richness(ps, measures=c("Shannon","Observed","Simpson"))

```


<br>

# **Beta Diversity**

```{r beta diversity}

# calculate the distance matrix with the transformed log data

ordinate<- ordinate(physeq=ps_log, method = "PCoA", distance = "bray")


# plot beta diversity
p2 <-
    plot_ordination(
        physeq = ps_log,
        ordination = ordinate,
        color = "Type",
        label = "Sample"
    ) + geom_point(size = 5) + theme_bw()

p2

# One way to show separation of Types is by adding polygons
p2 +  geom_polygon(aes(fill=Type))

```


We are stopping here for the sake of time but you can do a few more analysis with this data like differential abundance analysis.



